---
title: "The Relationship Between Life Expectancy and the Gini Coefficient"
author: 
  - "Vishwamber Reddy"
  - "Caoyu Shao"
  - "Hongyu Yin"
format:
  revealjs:
    theme: simple
    slide-number: true
    show-slide-number: all
    footer: "Life Expectancy and Income Inequality Analysis"
    scrollable: true
    smaller: true
editor: visual
execute:
  echo: false
  warning: false
---

## Introduction {.smaller}

-   Research Question: What is the relationship between income inequality and life expectancy?
-   Time period: 1990-2020
-   Global scope with country-level analysis

## Data Sources {.smaller}

-   Life expectancy data from Our World in Data
-   Gini coefficient data from World Bank
-   Coverage: 30-year period (1990-2020)
-   Key metrics:
    -   Life expectancy at birth
    -   Gini coefficient (0-1 scale)

## Methodology: {.smaller}

1.  Data Collection & Cleaning:

-   Imported life expectancy and Gini coefficient data from Our World in Data

-   Filtered data for 1990-2020 period

-   Merged datasets and removed NA values

2.  Analysis Steps:

-   Calculated summary statistics and correlation between variables

-   Created time trend analysis showing global averages over 30 years

-   Generated scatter plots by decades to visualize relationship changes

-   Compared 20 selected countries (10 high inequality vs 10 low inequality) based on mean values

## 

3.  Visualization Approach:

-   Used dual-axis plots to show both metrics simultaneously

-   Applied color coding to distinguish between high/low inequality countries

-   Implemented trend lines and point markers for clear pattern identification

```{r load-libraries-and-data}
# Load required libraries
library(tidyverse)      
library(knitr)         
library(ggplot2)       
library(dplyr)         
library(scales)        

# Fetch data directly from URLs
life_expectancy <- read.csv("https://ourworldindata.org/grapher/life-expectancy.csv?v=1&csvType=full&useColumnShortNames=false")

gini_index <- read.csv("https://ourworldindata.org/grapher/economic-inequality-gini-index.csv?v=1&csvType=full&useColumnShortNames=false")

cleaned_data <- life_expectancy %>%
  # Select relevant columns and rename for clarity
  select(Entity, Code, Year, 
         Life_Expectancy = Period.life.expectancy.at.birth...Sex..total...Age..0) %>%
  # Filter years between 1990 and 2020
  filter(Year >= 1990 & Year <= 2020) %>%
  # Join with gini data
  inner_join(
    gini_index %>%
      select(Entity, Code, Year, 
             Gini = Gini.coefficient..2017.prices....Income.or.consumption.consolidated) %>%
      filter(Year >= 1990 & Year <= 2020),
    by = c("Entity", "Code", "Year")
  ) %>%
  # Remove any remaining NA values
  na.omit()
```

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for life expectancy and Gini coefficient (1990-2020)"

# Calculate correlation
correlation <- cor(cleaned_data$Life_Expectancy, cleaned_data$Gini)

summary_stats <- cleaned_data %>%
  summarise(
    Mean_Life_Expectancy = mean(Life_Expectancy),
    SD_Life_Expectancy = sd(Life_Expectancy),
    Mean_Gini = mean(Gini),
    SD_Gini = sd(Gini)
  )
kable(summary_stats)

```

Table and Figure References:

As shown in @tbl-summary-statistics, the average life expectancy was 72.82 years (SD = 7.84), and the average Gini coefficient was 0.38 (SD = 0.09), reflecting substantial variability across countries.

## 

```{r}
#| label: fig-time-trends
#| fig-cap: "Global trends in life expectancy and Gini coefficient from 1990 to 2020"

# Time trend analysis
time_trend <- cleaned_data %>%
  group_by(Year) %>%
  summarize(
    mean_life_exp = mean(Life_Expectancy),
    mean_gini = mean(Gini)
  )

ggplot(time_trend, aes(x = Year)) +
  geom_line(aes(y = mean_life_exp, color = "Life Expectancy"), size = 1.2) +
  geom_line(aes(y = mean_gini * 100, color = "Gini (×100)"), size = 1.2) +  # scale Gini for visibility
  scale_y_continuous(
    name = "Life Expectancy",
    sec.axis = sec_axis(~ . / 100, name = "Gini Coefficient")
  ) +
  scale_color_manual(values = c("Life Expectancy" = "blue", "Gini (×100)" = "red")) +
  labs(
    title = "Trends in Life Expectancy and Income Inequality, 1990-2020",
    x = "Year",
    color = "Metric"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "red"),
    axis.title.y.left = element_text(color = "blue")
  )
```

@fig-time-trends illustrates trends over time, showing a steady increase in life expectancy and a mild decline in income inequality from 1990 to 2020.

## Results: {.smaller}

::: {style="font-size: 0.8em;"}
Our results demonstrate a negative correlation between income inequality and life expectancy (r = -0.45). Countries with higher Gini coefficients tend to have lower life expectancy.

Over the 30-year period, life expectancy increased globally from approximately 70 years in 1990 to 76 years in 2020. Meanwhile, income inequality slightly declined, with the average Gini coefficient decreasing from 0.39 to 0.35.

```{r}
#| label: fig-scatter-plot-decades
#| fig-cap: "Relationship between income inequality and life expectancy"

# Scatter plot with decades
cleaned_data_decades <- cleaned_data %>%
  filter(Year %in% c(1990, 2000, 2010, 2020))

ggplot(cleaned_data_decades, aes(x = Gini, y = Life_Expectancy, color = as.factor(Year))) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method="lm", color="blue") +
  scale_color_viridis_d() +  # Discrete color scale for decades
  labs(
    title = "Relationship Between Income Inequality and Life Expectancy by Decade",
    x = "Gini Coefficient",
    y = "Life Expectancy (years)",
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 12, face = "bold")
  )
```

@fig-scatter-plot-decades presents cross-sectional snapshots by decade. The scatter plots show a persistent negative relationship across time points, supporting the association between inequality and health outcomes.
:::

## 

::: {style="font-size: 0.8em;"}
Notably, countries with Gini coefficients below 0.3 generally achieved life expectancies above 80 years, while nations with coefficients above 0.45 frequently had life expectancies under 70 years.

The relationship remained consistent across decades, though the strength of the correlation varied by region and development level. This suggests that while income inequality is an important factor in life expectancy, other variables likely play significant roles in determining population health outcomes.
:::

```{r}
#| label: fig-main-visualization
#| fig-cap: "Comparison of life expectancy and income inequality"

# Check which countries have sufficient data
countries_with_sufficient_data <- cleaned_data %>%
  group_by(Entity) %>%
  summarize(
    count = n(),
    mean_life_exp = mean(Life_Expectancy, na.rm = TRUE),
    mean_gini = mean(Gini, na.rm = TRUE)
  ) %>%
  filter(count > 5, !is.na(mean_life_exp), !is.na(mean_gini))

# Get 10 countries with high Gini and low life expectancy
high_inequality_countries <- countries_with_sufficient_data %>%
  filter(mean_life_exp < 70) %>%
  arrange(desc(mean_gini), desc(-mean_life_exp)) %>%
  head(10)

# Get 10 countries with low Gini and high life expectancy
low_inequality_countries <- countries_with_sufficient_data %>%
  filter(mean_life_exp > 70) %>%
  arrange(mean_gini, desc(mean_life_exp)) %>%
  head(10)

# Combine the two groups
selected_countries <- bind_rows(
  high_inequality_countries %>% mutate(group = "High Inequality"),
  low_inequality_countries %>% mutate(group = "Low Inequality")
)

ggplot(selected_countries) +
  geom_bar(aes(x = reorder(Entity, mean_life_exp), y = mean_life_exp, fill = group), 
          stat = "identity", alpha = 0.7) +
  geom_point(aes(x = Entity, y = mean_gini * 100), color = "red", size = 3) +
  scale_y_continuous(
    name = "Life Expectancy (years)",
    sec.axis = sec_axis(~ . / 100, name = "Gini Coefficient")
  ) +
  scale_fill_manual(values = c("High Inequality" = "coral", "Low Inequality" = "skyblue")) +
  labs(
    title = "Life Expectancy and Income Inequality in Selected Countries",
    subtitle = "Comparison of Countries with High and Low Inequality (1990-2020)",
    x = "Country",
    fill = "Inequality Level"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "red"),
    axis.title.y.left = element_text(color = "blue"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12, face = "bold")
  )

```

## Conclusion {.smaller}

According to our discussion, we get a conclusion that there is a negative relationship between gini-coefficient and life-expectation. We get four explanations for that.

::: {style="font-size: 0.8em;"}
1.  Resource Inequality

-   Higher Gini coefficients indicate greater concentration of wealth among fewer people
-   Rich populations access better healthcare, preventive screenings, and insurance
-   Poor populations often delay or forgo treatment due to cost
-   Affluent households maintain healthier lifestyles through better nutrition and living conditions

2.  Psychosocial Stress and Social Capital

-   Large income gaps increase social comparison and status anxiety
-   Chronic stress responses harm cardiovascular, immune, and endocrine systems
-   High-inequality societies show reduced trust and mutual support
-   Weakened social networks limit health information sharing and community care
:::

------------------------------------------------------------------------

::: {style="font-size: 0.8em;"}
3.  Underinvestment in Public Health and Welfare

-   Unequal societies often underfund universal healthcare and public health programs
-   Limited welfare services reduce overall population longevity
-   Education gaps affect health literacy and preventive behaviors
-   Lower-educated groups face increased chronic disease risks

4.  Cumulative and Intergenerational Effects

-   Childhood poverty impacts physical and cognitive development
-   Early life disadvantages persist into adulthood
-   Limited social mobility in unequal societies
-   Health disadvantages perpetuate across generations
:::

##  {.center background-color="#f7f7f7"}

::: {.center style="font-size: 4em; text-align: center;"}
*Thank You*
:::
