---
title: "The Relationship Between Life Expectancy and the Gini Coefficient"
author: 
  - "Vishwamber Reddy"
  - "Cayou Shao"
  - "Hongyu Yin"
format:
  revealjs:
    theme: simple
    slide-number: true
    show-slide-number: all
    footer: "Life Expectancy and Income Inequality Analysis"
    scrollable: true
    smaller: true
editor: visual
execute:
  echo: false
  warning: false
---

## Introduction {.smaller}

-   Research Question: What is the relationship between income inequality and life expectancy?
-   Time period: 1990-2020
-   Global scope with country-level analysis

## Data Sources

-   Life expectancy data from Our World in Data
-   Gini coefficient data from World Bank
-   Coverage: 30-year period (1990-2020)
-   Key metrics:
    -   Life expectancy at birth
    -   Gini coefficient (0-1 scale)

## Methodology: {.smaller}

To investigate the relationship between income inequality and life expectancy, we analyzed data from two primary sources: life expectancy data from Our World in Data and Gini coefficient data from the World Bank. The analysis focused on the period from 1990 to 2020, covering multiple countries worldwide.

The Gini coefficient, ranging from 0 to 1, measures income inequality within countries, where 0 represents perfect equality and 1 represents maximum inequality. Life expectancy at birth represents the average number of years a newborn would live under current mortality rates.

```{r load-libraries-and-data}
# Load required libraries
library(tidyverse)      
library(knitr)         
library(ggplot2)       
library(dplyr)         
library(scales)        

# Fetch data directly from URLs
life_expectancy <- read.csv("https://ourworldindata.org/grapher/life-expectancy.csv?v=1&csvType=full&useColumnShortNames=false")

gini_index <- read.csv("https://ourworldindata.org/grapher/economic-inequality-gini-index.csv?v=1&csvType=full&useColumnShortNames=false")

cleaned_data <- life_expectancy %>%
  # Select relevant columns and rename for clarity
  select(Entity, Code, Year, 
         Life_Expectancy = Period.life.expectancy.at.birth...Sex..total...Age..0) %>%
  # Filter years between 1990 and 2020
  filter(Year >= 1990 & Year <= 2020) %>%
  # Join with gini data
  inner_join(
    gini_index %>%
      select(Entity, Code, Year, 
             Gini = Gini.coefficient..2017.prices....Income.or.consumption.consolidated) %>%
      filter(Year >= 1990 & Year <= 2020),
    by = c("Entity", "Code", "Year")
  ) %>%
  # Remove any remaining NA values
  na.omit()
```

We performed the following data processing and analysis steps:

1.  Data cleaning by removing missing values and standardizing country names

2.  Merging the datasets based on country and year

3.  Computing summary statistics and correlation analysis

4.  Creating visualizations to examine relationships and trends

The cross-sectional analysis examined the relationship between life expectancy and Gini coefficient at select years. The longitudinal analysis tracked changes in both variables across time.

## 

Table and Figure References:

As shown in Table 1, the average life expectancy was 72.82 years (SD = 7.84), and the average Gini coefficient was 0.38 (SD = 0.09), reflecting substantial variability across countries.

```{r summary-statistics}
#| tbl-cap: "Summary statistics for life expectancy and Gini coefficient (1990-2020)"

# Calculate correlation
correlation <- cor(cleaned_data$Life_Expectancy, cleaned_data$Gini)

summary_stats <- cleaned_data %>%
  summarise(
    Mean_Life_Expectancy = mean(Life_Expectancy),
    SD_Life_Expectancy = sd(Life_Expectancy),
    Mean_Gini = mean(Gini),
    SD_Gini = sd(Gini)
  )
kable(summary_stats)

```

Figure 1 illustrates trends over time, showing a steady increase in life expectancy and a mild decline in income inequality from 1990 to 2020.

```{r time-trend-analysis}
#| fig-cap: "Global trends in life expectancy and Gini coefficient from 1990 to 2020"

# Time trend analysis
time_trend <- cleaned_data %>%
  group_by(Year) %>%
  summarize(
    mean_life_exp = mean(Life_Expectancy),
    mean_gini = mean(Gini)
  )

ggplot(time_trend, aes(x = Year)) +
  geom_line(aes(y = mean_life_exp, color = "Life Expectancy"), size = 1.2) +
  geom_line(aes(y = mean_gini * 100, color = "Gini (×100)"), size = 1.2) +  # scale Gini for visibility
  scale_y_continuous(
    name = "Life Expectancy",
    sec.axis = sec_axis(~ . / 100, name = "Gini Coefficient")
  ) +
  scale_color_manual(values = c("Life Expectancy" = "blue", "Gini (×100)" = "red")) +
  labs(
    title = "Trends in Life Expectancy and Income Inequality, 1990-2020",
    x = "Year",
    color = "Metric"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "red"),
    axis.title.y.left = element_text(color = "blue")
  )
```

## Results: {.smaller}

Our results demonstrate a negative correlation between income inequality and life expectancy (r = -0.45). Countries with higher Gini coefficients tend to have lower life expectancy.

Over the 30-year period, life expectancy increased globally from approximately 70 years in 1990 to 76 years in 2020. Meanwhile, income inequality slightly declined, with the average Gini coefficient decreasing from 0.39 to 0.35.

Figure 2 presents cross-sectional snapshots by decade. The scatter plots show a persistent negative relationship across time points, supporting the association between inequality and health outcomes.

```{r scatter-plot-decades}
#| fig-cap: "Relationship between income inequality and life expectancy"

# Scatter plot with decades
cleaned_data_decades <- cleaned_data %>%
  filter(Year %in% c(1990, 2000, 2010, 2020))

ggplot(cleaned_data_decades, aes(x = Gini, y = Life_Expectancy, color = as.factor(Year))) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method="lm", color="blue") +
  scale_color_viridis_d() +  # Discrete color scale for decades
  labs(
    title = "Relationship Between Income Inequality and Life Expectancy by Decade",
    x = "Gini Coefficient",
    y = "Life Expectancy (years)",
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 12, face = "bold")
  )
```

## 

Notably, countries with Gini coefficients below 0.3 generally achieved life expectancies above 80 years, while nations with coefficients above 0.45 frequently had life expectancies under 70 years.

The relationship remained consistent across decades, though the strength of the correlation varied by region and development level. This suggests that while income inequality is an important factor in life expectancy, other variables likely play significant roles in determining population health outcomes.

```{r main-visualization}
#| fig-cap: "Comparison of life expectancy and income inequality"

# Check which countries have sufficient data
countries_with_sufficient_data <- cleaned_data %>%
  group_by(Entity) %>%
  summarize(
    count = n(),
    mean_life_exp = mean(Life_Expectancy, na.rm = TRUE),
    mean_gini = mean(Gini, na.rm = TRUE)
  ) %>%
  filter(count > 5, !is.na(mean_life_exp), !is.na(mean_gini))

# Get 10 countries with high Gini and low life expectancy
high_inequality_countries <- countries_with_sufficient_data %>%
  filter(mean_life_exp < 70) %>%
  arrange(desc(mean_gini), desc(-mean_life_exp)) %>%
  head(10)

# Get 10 countries with low Gini and high life expectancy
low_inequality_countries <- countries_with_sufficient_data %>%
  filter(mean_life_exp > 70) %>%
  arrange(mean_gini, desc(mean_life_exp)) %>%
  head(10)

# Combine the two groups
selected_countries <- bind_rows(
  high_inequality_countries %>% mutate(group = "High Inequality"),
  low_inequality_countries %>% mutate(group = "Low Inequality")
)

ggplot(selected_countries) +
  geom_bar(aes(x = reorder(Entity, mean_life_exp), y = mean_life_exp, fill = group), 
          stat = "identity", alpha = 0.7) +
  geom_point(aes(x = Entity, y = mean_gini * 100), color = "red", size = 3) +
  scale_y_continuous(
    name = "Life Expectancy (years)",
    sec.axis = sec_axis(~ . / 100, name = "Gini Coefficient")
  ) +
  scale_fill_manual(values = c("High Inequality" = "coral", "Low Inequality" = "skyblue")) +
  labs(
    title = "Life Expectancy and Income Inequality in Selected Countries",
    subtitle = "Comparison of Countries with High and Low Inequality (1990-2020)",
    x = "Country",
    fill = "Inequality Level"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "red"),
    axis.title.y.left = element_text(color = "blue"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12, face = "bold")
  )

```

## Conclusion

According to our discussion, we get a conclusion that there is a negative relationship between gini-coefficient and life-expectation. We get four explanations for that.

First, due to resource inequality, there will be more poor people and less rich people if there is a high gini coefficient. Rich people could afford higher-quality care, preventive screenings, and insurance, while poorer people may delay or forgo treatment due to cost. And besides Affluent households can buy healthier food and live in cleaner, less crowded environments; low-income families face greater exposure to pollution, overcrowding, and food insecurity.

Second, due to Psychosocial Stress and Social Capital, large income gaps fuel social comparison and status anxiety, triggering persistent stress responses that harm cardiovascular, immune, and endocrine systems. What's more, High-inequality societies often exhibit lower trust and mutual support, reducing the informal networks that share health information and provide neighborhood‐level care.

---

Third, due to underinvestment in Public Health and Welfare.When tax, welfare, and public services disproportionately benefit the well-off, funding for universal healthcare, elder care, and child health programs suffers—limiting gains in overall longevity.Besides, Education strongly influences health literacy and behaviors; in unequal societies, lower-educated groups lack knowledge about disease prevention and healthy lifestyles, increasing chronic‐disease risks.

Lastly,due to cumulative and Intergenerational Effect, Childhood poverty—malnutrition, exposure to toxins, or family financial stress—can stunt physical and cognitive development, with effects that persist into adulthood.Besides, in highly unequal settings, children from poor families have fewer opportunities to escape poverty, perpetuating health disadvantages across generations.
